Efficient Calculation of Polynomial and Interaction Features on Sparse Matrices

### 2015-10-06 paper/paper2.tex
Added description of index-mapping
Added description of index mapping and its derivation, and
generalization to interactions among 3 features.

What is therefore needed is a mapping between column index pairs of $\vec{x}$ and the space ${0, 1, ..., \frac{D^2+D}{2}-1}$
(the output size of second degree polynomial features when $D = |\vec{x}|$). The ordering of the mapping is irrelevant so long as its input

@@

To construct this mapping, consider a matrix of size $D \times D$ where cell $(a,b)$ represents the
polynomial feature between indices $a$ and $b$ for $a,b \in I$ for vector $\vec{x}$.

\renewcommand{\kbldelim}{(}% Left delimiter
\renewcommand{\kbrdelim}{)}% Right delimiter
\[
  \kbordermatrix{
             & x_0    & x_1   & \cdot   & \cdot   &  \cdot  &  x_{D-1} \\
     x_0     & 0      & 1     & \cdot   &  \cdot  &  \cdot  &  D-1     \\
     x_1     & -      & D     & \cdot   &  \cdot  &  \cdot  &  2D-2    \\
     \cdot   & \cdot  & \cdot &  \cdot  &      &      &  \cdot       \\
     \cdot   & \cdot  &       &  \cdot  &  \cdot  &      &  \cdot       \\
     \cdot   & \cdot  &       &         &  \cdot  &  \cdot  &  \cdot       \\
    x_{D-1}  & -      & \cdot &  \cdot  & \cdot   & -    &  \frac{D^2+D}{2}-1\\
  }
\]

Again, any bijective mapping $I \mapsto O$ would suffice, but this particular mapping could
be considered one of the more natural options as it maps $I$ onto $O$ where both $I$ and $O$ are sorted in ascending order.
As can be seen, the required mapping can be viewed as mapping the upper triangular matrix (including 
the diagonal), whose indicies are $I$, onto $O$. The construction of this mapping will now
be given.

INSERT JOHN'S PROOF HERE

The mapping can be modified slightly to yield interaction features instead of polynomial features
by noting that the section of the $D \times D$ mapping matrix that will yield interaction
features differs only in that its diagonal is excluded. The mapping is therefore gotten by
taking the polinomial mapping and subtracting $i+1$.

>>>>

What is therefore needed is a mapping between row-column index pairs of $\vec{x}$ and the space ${0, 1, ..., \frac{D^2+D}{2}-1}$
(the output size of second degree polynomial features when $D = |\vec{x}|$). The ordering of the mapping is irrelevant so long as its input

@@

Interaction features are a way of capturing correlations between features in a machine 
learning setting. A feature vector $\vec{x}$ of dimensionality $D$ has second degree interaction features 
$\{x_i \cdot x_j : i, j \in \{0,1,..., D-1\} \land i < j\}$, 
so a $D$ dimensional vector has $\binom{D}{2} = \frac{D^2-D}{2}$ second degree interaction features. A naive
approach to calculating these features is to simply iterate through the combinations of the column indices.
For a sparse vector, many of the resulting interaction features would be zero, and could therefore be ignored.
This work describes a method to efficiently calculate second degree interaction features for a sparse matrix 
that has time and space complexities that decrease quadratically with the density of the input matrix with respect to the naive approach.

\section{Approach}
Let the list of nonzero columns for a given row $\vec{x}$ be denoted by $N_{zc}$. The nonzero second degree 
interaction features are simply the products of all combinations of two elements whose 
columns are in $N_{zc}$. However, to properly place an interaction feature into the correct column, a mapping from the column 
index pairs of $N_{zc}$ into the columns of the interaction matrix is needed. The mapping is 
from pairs $(a, b)$,  where $a$ and $b$ are in $1,2,..., D$, and $a < b$, to $1,2,..., \frac{D^2-D}{2}$,  Such a mapping essentially consists of mapping the indices of entries in the upper triangle of a matrix to indices in a flat 
list. We now describe the construction of such a mapping.

\subsection{Mapping construction}
\textbf{TODO: convert all $i$ and $j$ indices into $a$ and $b$.--jfh}

We seek a map from matrix indices $(i, j)$ (with $i < j$ and $0 \le i < D$) to numbers $f(i, j)$ with $0 \le f(i, j) < \frac{D(D-1)}{2}$, one that follows the pattern indicated by 
\begin{align}
\begin{bmatrix}
x & 0 & 1 & 3 \\
x & x & 2 & 4 \\
x & x & x & 5 \\
x & x & x & x
\end{bmatrix}
\label{eq:4x4mat}
\end{align}
where the entry at location $(i,j)$ displays the value $f(i, j)$.

To simplify slightly, we introduce a notation for the $n$th triangular number, 
\begin{align}
T_2(n) = \frac{n(n+1)}{2}
\end{align}
\noindent
The subscript $2$ is there to indicate that these are triangles in two dimensions; we'll use $T_3(n)$ to indicate the $n$th tetrahedral number, and so on for higher dimensions.

Observe that in Equation~\ref{eq:4x4mat}, each entry in column $j$ lies in the range
\begin{align}
T_2(j-1) &\le e < T_2(j).
\end{align}
\noindent
And in fact, the entry in the $i$th row of that column is just $i + T_2(j-1)$. Thus we have
\begin{align}
f(i, j) 
&= i + T_2(j-1)\\
&= i + \frac{(j-1)j}{2}\\
&=  \frac{2i + j^2-j}{2}.
\end{align}

For instance, in column $j = 2$ in our example (the \emph{third} column), the entries range from $1$ to $2$, while $T_2(j-1) = T_2(1) = 1$ and $T_2(j) = T_2(2) = 3$, and the entry in column $j = 2$, row $i = 1$ is 
$i + T_2(j-1) = 1 + 1 = 2$.

\subsubsection{Other indices}
With one-based indexing in both the domain and codomain, the formula above becomes
\begin{align}
f_1(i, j) &= 1+ f(i-1, j-1) \\
&= 1+ f(i-1, j-1) \\
& = \frac{2 + 2(i-1) + (j-1)^2-(j-1)}{2}\\
& = \frac{2i + j^2 - 3j + 2}{2}
\end{align}
\subsubsection{Higher dimensions}
To handle three-way interactions, we need to map triples of indices in a 3-index array to a flat list, and similarly for higher-order interactions.

For three indices, $i,j,k$, with $i < j < k$ and $0 \le i,j,k < D$, we have a similar recurrence. Calling the mapping $g$, we have 
\begin{align}
g(i,j,k) = i + T_2(j-1) + T_3(k-1);
\end{align}
if we define $T_1(i) = i$, then this has the very regular form
\begin{align}
g(i,j,k) = 1 + T_1(i-1) + T_2(j-1) + T_3(k-1);
\end{align}
and from this, the generalization to higher dimensions is straightforward. The formulas for ``higher triangular numbers'', i.e., those defined by
\begin{align}
T_k(n) &= \sum_{i=1}^n T_{k-1}(n)
\end{align}
for $k > 1$ can be determined inductively. For $k = 3$, the result is 
\begin{align}
T_3(n) &= \sum_{i=1}^n T_{2}(n)\\
&= \frac{n^3 + 3n^2 + 2n}{6},
\end{align}
so that the formula for 3-way interactions, with zero-based indexing, becomes 
\begin{align}
g(i, j, k) &= 1 + (i-1) + \frac{(j-1)j}{2} + \\
& \frac{(k-1)^3 + 3(k-1)^2 + 2(k-1)}{6}. 
\end{align}
\subsection{Inversion}
We not only want to be able to compute the ``flat index'' $f(i,j)$ corresponding to a particular index pair $(i, j)$, we'd like, given a value $N = f(i, j)$, to be able to determine $i$ and $j$. We know, given $N$, that $f(0, j) \le N \le f(j,j-1)$, i.e., that 
\begin{align}
\frac{j^2 - 3j + 2}{2} &\le N &\le \frac{2(j-1) + j^2 - 3j + 2}{2}\\
{j^2 - 3j + 2} &\le 2N &\le {j^2 - j}.
\end{align}
If we find the (real number) value $j_{*}$ for which $j_{*}^2 - j_{*} = 2N$, then we know that $j = \lfloor j_{*}\rfloor$. Finding $j_{*}$ amounts to solving a quadratic, whence
\begin{align}
j_{*} = \frac{1 \pm \sqrt{1 + 8N}}{2}.
\end{align}
Clearly the larger value, 
\begin{align}
j_{*} = \frac{1 + \sqrt{1 + 8N}}{2},
\end{align}
is the one we seek; we set $j = \lfloor j_{*}\rfloor$, and compute $i = N - \frac{j^2 - j}{2}$.

In practice, repeatedly computing square roots may be annoying; instead, when we compute $N = f(i,j)$, we can record the values of $i$ and $j$ along with the interaction features, making inversion a simple matter of looking up these values.

%To construct this mapping, consider a matrix of size $D \times D$ where %cell $(a,b)$ represents the
%polynomial feature between indices $a$ and $b$ for $a,b \in I$ for vector $\vec{x}$.

%\renewcommand{\kbldelim}{(}% Left delimiter
%\renewcommand{\kbrdelim}{)}% Right delimiter
%\[
%  \kbordermatrix{
%             & x_0    & x_1   & \cdot   & \cdot   &  \cdot  &  x_{D-1} \\
%     x_0     & 0      & 1     & \cdot   &  \cdot  &  \cdot  &  D-1     \\
%     x_1     & -      & D     & \cdot   &  \cdot  &  \cdot  &  2D-2    \\
%     \cdot   & \cdot  & \cdot &  \cdot  &      &      &  \cdot       \\
%     \cdot   & \cdot  &       &  \cdot  &  \cdot  &      &  \cdot       %\\
%     \cdot   & \cdot  &       &         &  \cdot  &  \cdot  &  \cdot       %\\
%    x_{D-1}  & -      & \cdot &  \cdot  & \cdot   & -    &  \frac{D^2+D}%{2}-1\\
  %}
%\]

%Again, any bijective mapping $I \mapsto O$ would suffice, but this %particular mapping could
%be considered one of the more natural options as it maps $I$ onto $O$ %where both $I$ and $O$ are sorted in ascending order.
%As can be seen, the required mapping can be viewed as mapping the upper %triangular matrix (including 
%the diagonal), whose indicies are $I$, onto $O$. The construction of this %mapping will now
%be given.

%The mapping can be modified slightly to yield interaction features %instead of polynomial features
%by noting that the section of the $D \times D$ mapping matrix that will %yield interaction
%features differs only in that its diagonal is excluded. The mapping is %therefore gotten by
%taking the polinomial mapping and subtracting $i+1$.
